---
title: "Error Mean function 01"
author: "Tahmidul Islam"
date: "7/1/2020"
output: 
 html_document:
  fig_width: 14 
  fig_height: 6
---

```{r setup, include = FALSE, echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE, autodep = TRUE, error = TRUE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = TRUE, autodep = TRUE, error = TRUE)
```

```{r}
#loading packages
rm(list = ls())
library(MASS)
library(kernlab)
library(mvtnorm)
library(Matrix)
library(optimx)
library(fdapace)
library(dplyr)
library(cubature)
library(smoothmest)
library(kableExtra)
library(colorspace)

source('../code/RBF.R')
source('../code/functionsTest.R')
#source('../code/functionsNew.R')
#source('code/laplace.R')
#source('code/matern52.R')
#source('code/matern32.R')
source('../code/multistart.R')
```

## Mean Integrated Squared Error (MISE)

The mean integrated squared error (MISE) is defined by
$$
E || \hat f - f ||^2 = E \int \big( \hat f(t) - f(t) \big)^2 dt.
$$

Where $f$ is the true function. We first compute the integrated squared error.

```{r, results = 'markup', echo = TRUE}
# Compute error
source('../code/RBF.R')
rint <- integrate(residfunc, lower = 0 , upper = 1, muf = muf1, est = gpsmooth, 
          est_arg = fet.muf1.rbf, subdivisions = 10000)$value
source('../code/laplace.R')
rint <- cbind(rint,
integrate(residfunc, lower = 0 , upper = 1, muf = muf1, est = gpsmooth, 
          est_arg = fet.muf1.lap, subdivisions = 10000)$value)
source('../code/matern52.R')
rint <- cbind(rint,
integrate(residfunc, lower = 0 , upper = 1, muf = muf1, est = gpsmooth, 
          est_arg = fet.muf1.m52, subdivisions = 10000)$value)
source('../code/matern32.R')
rint <- cbind(rint,
integrate(residfunc, lower = 0 , upper = 1, muf = muf1, est = gpsmooth, 
          est_arg = fet.muf1.m32, subdivisions = 10000)$value)
rint <- cbind(rint,
integrate(residfunc, lower = 0, upper = 1, muf = muf1, est = fpcamu, 
          est_arg = fpca1, subdivisions = 10000)$value)
```

Since R's base code for integration using adaptive quadrature sometimes results in error, we use a package called 'cubature'. This package uses adative and monte-carlo integration. We show results for the mix of normal density mean function.

```{r}
source('../code/RBF.R')
cubint <- cubintegrate(residfunc, lower = 0 , upper = 1, method = "pcubature", muf = muf1, est = gpsmooth, est_arg = fet.muf1.rbf)$integral
source('../code/laplace.R')
cubint <- cbind(cubint,
cubintegrate(residfunc, lower = 0 , upper = 1, method = "pcubature", muf = muf1, est = gpsmooth, est_arg = fet.muf1.lap)$integral)
source('../code/matern52.R')
cubint <- cbind(cubint,
cubintegrate(residfunc, lower = 0 , upper = 1, method = "pcubature", muf = muf1, est = gpsmooth, est_arg = fet.muf1.m52)$integral)
source('../code/matern32.R')
cubint <- cbind(cubint,
cubintegrate(residfunc, lower = 0 , upper = 1, method = "pcubature", muf = muf1, est = gpsmooth, est_arg = fet.muf1.m32)$integral)
cubint <- cbind(cubint,
cubintegrate(residfunc, lower = 0 , upper = 1, method = "pcubature", muf = muf1, est = fpcamu, est_arg = fpca1)$integral)
```

```{r}
#rint <- cbind(rint, NA)
intresult <- round(rbind(rint, cubint), 4)
intresult <- cbind(c('Base R', 'Cubature'), intresult)
colnames(intresult) <- c('Method','Gaussian', 'Laplace', 'Matern_5/2', 'Matern 3/2', 'PACE')
intresult %>% kable() %>% kable_styling()
```

Clearly, there are discrepencies in the integration result. We implement a simple Monte Carolo integration using uniform distribution. We generate $N$ time points from uniform(-1,1) and approximate the integral.

$$
\int_a^b f(t) dt = (b-a) \frac{1}{N} \sum f(t_i).  
$$

```{r}
source('../code/RBF.R')
ise.muf1 <- vrmse(n, method = 'mc', muf = muf1, est = gpsmooth, est_arg = fet.muf1.rbf) 
source('../code/laplace.R')
ise.muf1 <- cbind(ise.muf1, vrmse(n, method = 'mc', muf = muf1, est = gpsmooth, est_arg = fet.muf1.lap)) 
source('../code/matern52.R')
ise.muf1 <- cbind(ise.muf1, vrmse(n, method = 'mc', muf = muf1, est = gpsmooth, est_arg = fet.muf1.m52)) 
source('../code/matern32.R')
ise.muf1 <- cbind(ise.muf1, vrmse(n, method = 'mc', muf = muf1, est = gpsmooth, est_arg = fet.muf1.m32)) 

ise.muf1 <- cbind(ise.muf1, vrmse(n, method = 'mc', muf = muf1, est = fpcamu, est_arg = fpca1)) 
ise.muf1 <- as.data.frame(ise.muf1)
```


```{r}
names(ise.muf1) <- kername
ise.muf1 <-cbind(n, ise.muf1)
ise.muf1 %>% kable() %>% kable_styling()
```

```{r}
plot(n, ise.muf1[,2], col = 1, pch = 16, ylim = c(min(ise.muf1[,-1]), max(ise.muf1[,-1])), type = 'l', 
     xlab = 'Grid size', ylab = 'ISE', lwd = 3, main = 'ISE')
for (i in 3:6) {
  lines(n, ise.muf1[,i], col = i-1, pch = 16, lwd = 3)
}
legend('bottom', c('Gaussian', 'Laplace', 'Matern_5/2', 'Matern 3/2', 'PACE'), lty = 1, col = 1:5, bty = 'n', horiz = T)
```